raptr: Representative and Adequate Prioritisation Toolkit in R

### 2016-10-12 code/rmarkdown/article.Rmd
+ actually check new version of article into vc

1. An underlying aim in conservation planning is to maximise the long-term persistence of biodiversity. To fulfill this aim, the processes that sustain biodiversity must be conserved. Many biodiversity processes can be preserved by conserving a representative sample of each biodiversity feature (species or ecosystem). This idea has for decades been encapsulated in the principle of \"representativeness\". For example, representing the range of genetic diversity or ability to persist in a range of climatic conditions might be crucial for the long-term persistence of a species in a world where environmental change is accelerating. Yet current approaches overwhelmingly focus on \"adequacy\", the concept of ensuring a target proportion of the distribution of a biodiversity feature is reserved.

2. To address this issue, we developed the \\texttt{raptr: Representative and Adequate Prioritisation Toolkit in R} to guide reserve selection using conventional area protection targets as well as explicit targets for representing intra-specific biodiversity processes. Users set \"amount targets\" to preserve an adequate proportion of habitat for each species. They also set \"representation targets\" to secure a representative sample of variation within features reflect a biodiversity process (eg. genetic characteristics between individuals in areas, or variation in climatic conditions between inhabited areas). We explored the functionality of this package using a simulated study and two case studies. In each study, we generated solutions that secured a proportion of the species' distributions using amount targets--representing conventional reserve selection--and compared them with solutions generated using both amount and space targets.

3. We show that explicitly considering biodiversity processes in reserve selection can substantially change the solution. The simulations suggest that including representativeness in conservation planning is important where biodiversity features have multi-modal distributions in an attribute space. The case studies show that explicitly considering representation-based targets for biodiversity processes can return solutions that could be far more effective in achieving long term biodiversity persistence and do not require much more area to be reserved than traditional approaches.

@@

<!--  title page -->
Short running title: Representative and adequate prioritisations \newline
Abstract word count: 348 / 350  
Total word count: 5481 / 7000  
Number of references: 37  
Number of figures: 8  
Number of tables: 0  
Number of textboxes: 0  
Number of equations: 8  
\newpage

# Introduction
<!-- conservation and reserve selection -->
Perhaps the most fundamental aim of conservation is to maximise the long-term persistence of biodiversity [@r423; @r255]. To achieve this, conservation actions must preserve biodiversity patterns (eg. populations, species, ecosystems), but also crucially the processes that sustain them. One of the major tangible achievements of modern conservation has been the act of setting aside areas for preservation [@r440]. Reserve networks buffer species from gross threatening processes and set the stage for enhanced management interventions [@r465]. Since the resources available for conservation action are limited, protected area networks must be sited in places that satisfy conservation objectives for minimal cost [@r255]. However, despite the importance of biodiversity processes, current approaches for incorporating them into reserve selection are limited.

<!-- biodiversity processes are important -->
Prioritisations must capture the ecological and evolutionary processes that sustain biodiversity [@r3; @r255; @r474]. Ecological processes are important for biodiversity to persist over short time-scales. For instance, predator-prey processes can mediate ecosystem structure and function through top-down control [@rXXX]. Additionally, migratory processes  allow many species to access suitable habitat for breeding [@rXXX]. On the other hand, evolutionary processes are important for biodiversity to persist over long time-scales. Adaptive evolutionary processes foster resilience against environmental change [@r3; @r43], and neutral evolutionary processes maintain genetic diversity [@r5; @r67; @r63]. In recent decades, a wealth of data relevant to biodiversity processes has become available to conservation planners [eg. bioclimatic, distribution, genetic, phylogenetic, and trait data; @rXXX; @r33; @r460; @r453]. This boom has resulted in significant progress towards incorporating both ecological and evolutionary processes into conservation planning [@r473; @r474; @r448; @r2; @r68; @r470; @r469; @r247; @r488; @r468; @r407].

<!-- current methods for preserving biodiversity processes -->
Many decision support tools have been developed to help identify cost-effective solutions [eg. \texttt{ConsNet}, @r446; \texttt{Marxan}, @r22; \texttt{Zonation}, @r429]. Typically, these tools are used to deliver a prioritisation that secures a set of species of interest to the conservation planner (termed target species or features). Given a set of species, decision makers use these tools to generate a prioritisation that secures an adequate proportion of each species' range. To accommodate biodiversity processes, current methods involve partitioning species' ranges into different groups as a pre-processing step to render a new set of "pseudo-species" [eg. at habitat discontinuities; @r2]. However, this approach suffers two main limitations. First, splitting species' ranges into discrete groups is not a trivial task. Data on biodiversity processes are often continuous, hierarchical, and hyper-dimensional, and so may not be able to be reduced to a few groups without significant information loss [@r218]. Second, by splitting the proxy variable into groups, the decision maker cannot directly assess how well a prioritisation actually samples variation in the original proxy variable. Simply setting targets for representing each of the pseudo-species does not necessarily guarantee that the original proxy variable been well sampled--and by extension that the biodiversity process has been preserved.

<!-- aims of this paper -->
One of the key issues in reserve selection is the lack of a unifying decision support tool that can accommodate data on biodiversity patterns and processes. To begin to fill this gap, we present the \texttt{raptr R} package. This \texttt{R} package uses novel formulations of the reserve selection problem to provide decision makers with the tools to generate prioritisations that preserve both biodiversity patterns and processes. Here, we aim to define the concepts that underpin this \texttt{R} package and explore its functionality.

# Methods
## PROBLEM FORMULATION
Biodiversity features are defined as the entities that the prioritisation is required to preserve (eg. species, ecosystems). Spatial attributes are defined as the variation across the species' range that the prioritisation is required to sample. They can be intrinsic (eg. genetic or trait variation) or extrinsic (eg. environmental variation) to the feature. These attributes should be related to the biodiversity processes that the decision maker aims to preserve.

Each attribute is conceptualised as a space. This space is termed an "attribute space". Each planning unit is thought to occupy a single point inside each space. For example, a decision maker may require a prioritisation that represents populations along climatic gradients. To achieve this, the decision maker might use an "climatic" attribute space with dimensions relating to mean annual temperature ($^{\circ}$C) and precipitation (mm). Any given combination of temperature and precipitation may be conceived as a point in this environmental space. By associating planning units with climatic data, they can be mapped from geographic space to this environmental attribute space.

Demand points are points that also exist in an attribute space. They are designated by the decision maker to indicate regions of the attribute space that the decision maker wishes to represent in the prioritisation (see below for discussion on how demand points can be generated for real-world datasets). The amount of variation in the attribute space that a prioritisation secures is a function of the distance between each demand point and each selected planning unit in the attribute space. The shorter the distances between the demand points and the planning units; the better the prioritisation is at securing the variation in the spatial attribute. To convert these amounts to a proportion--a meaningful unit for a decision maker--the distances between the selected planing units and the demand points are scaled by the distances between the demand points to the centroid of the demand points. In any attribute space there may exist points that are impossible (eg. mean annual rainfall -5 mm), or do not occur in the study area (eg. mean annual temperature 30$^{\circ}$C in Antarctica). Additionally, there may be some regions that are desirable for some features and undesirable for others (eg. conditions known to be outside the physiological tolerance of certain species). Thus a different set of demand points and weights are used for each attribute space and each feature. By placing demand points in desirable regions of an attribute space for a given feature, the decision maker can ensure that prioritisations secure the feature in planning units with spatial attributes that are desirable for that feature.

To illustrate these concepts, consider an example conservation planning scenario example involving an attribute space and demand points. We wish to develop a prioritisation for a single species that has four populations in the study area. However, we can only afford to preserve three populations. We aim preserve the adaptive landscape of the species, and to achieve this, we will preserve populations inhabiting different environmental conditions. To describe environmental variation, we obtained data on the environmental conditions (rainfall (mm) and temperature ($^{\circ}$C)) where each population is found. These environmental data are then used to construct a two-dimensional environmental attribute space. Next, we generated demand points as equi-distant points between the range of values where the populations were found. By comparing the distribution of the demand points to the distribution of the populations in the attribute space, we can identify a suitable prioritisation (Fig. 1). We can see that preserving both populations $A$ and $C$, effectively "doubles-up" on the same environmental characteristics, resulting in considerable redundancy in the solution. Instead, a more representative sample of the intra-specific variation could be captured by securing populations $A$, $B$, and $D$. This example demonstrates how the inclusion of biodiversity processes can guide the reserve selection process.

The formulations used to express the reserve selection problem in the \texttt{raptr R} package are based on a combination of the \texttt{Marxan} reserve selection problem and the uncapacitated facility location problems [@r451]. Although the \texttt{raptr R} package utlises two novel formulations, for brevity, we will define the simpler formulation--referred to as the unreliable formulation--below and define the more complex version--the reliable formulation--in the Supporting Information S1. These formulations are named after the unreliable and reliable facility location formulations from which they are based upon [@r16]. The key difference between these two formulations is that the reliable formulation explicitly considers the probability that planning units are occupied when calculating the proportion of an attribute space sampled in a solution. All mathematical terms defined hereafter are described in Table S1. For convenience, the cardinality of sets will be denoted using the same symbol used to denote the variable.

@@

Targets are used to ensure that prioritisations adequately preserve each species. Amount-based targets are used to ensure that the total amount of habitat preserved is sufficient. Let $T_f$ denote the expected amount of area that needs to be preserved for each feature $f \in F$. Space-based targets ensure that a sufficient proportion of the intra-specific variation is secured. Let $\tau_{fs}$ denote the space-based targets for feature $f \in F$ and attribute space $a \in A$. For convenience, these both types of targets are expressed as proportions in the \texttt{R} package.

@@

\begin{align*}
T_s &= \text{amount target for feature $f$} \tag*{eqn 1a}\\
%
\tau_{sa} &= \text{representation target for feature $f$ in attribute space $a$} \tag*{eqn 1b}\\
%
B &= \text{boundary length modifier (BLM): penalise fragmented solutions} \tag*{eqn 1c}\\
\end{align*}

@@

\begin{align*}
X_j
	&= \begin{cases}
		1, & \parbox{25em}{if planning unit $j$ is selected for conservation action} \tag*{eqn 2a} \\
		0, & \parbox{25em}{otherwise} \\
	\end{cases} \\
%
Y_{fsij} &= \begin{cases}
		1, & \parbox{25em}{if demand point $i$ is assigned to planning unit $j$ for feature $f$ in space $s$. } \tag*{eqn 2b} \\
		0, & \parbox{25em}{otherwise} \\
	\end{cases} \\
\end{align*}

@@

## OPTIMISATION
The unreliable formulation is non-linear. However, the non-linear components can be linearised using existing techniques. The expression $X_j X_k$ in (eqn 3a) can be linearised using methods described by Beyer et al. [-@r426]. Linearised versions of the problems can be solved using commercial exact algorithm solvers. The \texttt{raptr} R package provides functions to express conservation planning data as optimisation problems using linearised versions of the unreliable and reliable formulations. These optimisation problems can then be solved to generate prioritisations using the commercial \texttt{Gurobi} software suite (\url{http://www.gurobi.com}). Presently, academics can obtain a license at no cost from the Gurobi website.

@@

## SIMULATION STUDY
### Methods
We simulated a hypothetical study area with planning units arranged in $`r sqrt(sim.params.LST[[MODE]]$number.planning.units)` \times `r sqrt(sim.params.LST[[MODE]]$number.planning.units)`$ square grid (Fig. 2). We then simulated three species across this study area. The first species was simulated to represent a hyper-generalist--occurring in all planning units with equal probability (Fig. 2a). This species' distribution was based on a uniform distribution (eqn 8a). The second species was simulated to represent a species with an idealised distribution (Fig. 2b). It has a core range area, and is less likely to be found in areas that are distant from the core area. This species' distribution was simulated using the probability density function of a single multivariate normal distribution (eqn 8c). The third species represents a species with two distinct populations (Fig. 2c). This species' distribution was simulated to depict a bimodal distribution, based on the combination of the probability density functions of two multivariate normal distributions (eqn 8d). The probability that each species inhabited a given planning unit was calculated using the $X, Y$ coordinates of the planning unit and eqns 8a--8d. We used a geographic attribute space to showcase the behavior of the problem. For each species, demand points were generated by calculating the centroids of each planning units and their weights were set as the probability that the planning units were occupied.

@@

The uniform species was simulated to occur with a constant probability of occupancy across the study area (Fig. 2a). The solution generated using `r sim.params.LST[[MODE]]$amount.target*100` % amount-based targets (Fig. 3a) selected `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units near the southern end of study area. This configuration is an artifact of the method used to solve this particular instance of the reserve selection problem. In fact, for this species, all prioritisations containing `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units are optimal when considering only `r sim.params.LST[[MODE]]$amount.target*100` % amount targets. In the absence of criteria to guarantee representativeness, reserve selection methods may or may not return solutions that secure a representative sample of the features. In this particular case, the solution does not secure a representative sample of the species' range (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$space.held,2)` % sampled). Note that the proportion here is negative because it performs worse than simply citing a planning unit in the centroid of the species' demand points.

@@

@@

By using `r sim.params.LST[[MODE]]$amount.target*100` % amount targets and `r sim.params.LST[[MODE]]$space.target*100` % geographic targets and a boundary length modifier parameter ($BLM = 1$), we obtained a solution that fulfills adequacy, representativeness and connectivity objectives (Fig. 3h). This solution contains `r filter(sim.spp.DF, species=='Normal\nspecies', prioritisation=='Amount & space\ntargets and BLM')$n` planning units. Similar to the corresponding solution for the uniform species (Fig. 3d), this solution contains more planning units than any other solutions for this species (cf. Figs. 3e--3g). The results for the normally distributed species suggest that the space targets can result in solutions that secure a more representative sample of the species'--even for species without significant structure.

@@

## CASE STUDY 1
### Methods
We investigated how space-based targets can be used in a multi-species planning context to generate a prioritisation that sufficiently preserves the realised niche for several species. By preserving the populations in suitable habitats with different with environmental conditions, conservation planners can preserve the species' adaptive landscape and foster resilience against environmental change [@r63]. We selected Queensland, Australia as the study area. This region is ideal for exploring the potential of niche-based targets because it contains a broad range of different habitats. We obtained data for 19 bioclimatic variables across the region [at $30^{\prime \prime}$ resolution from \url{www.worldclim.org}; @r33] and subjected them to a principal components analysis (using ArcMap 10.3.1). We used the first `r as.character(as.english(cs1.params.LST[[MODE]]$n.pc.layers))` principal components (cumulatively explained `r round(cs1.pca.DF[[4]][cs1.params.LST[[MODE]]$n.pc.layers],1)` % of the total variation) to characterise the environmental variation across the study area (Fig. 4).

We selected four bird species that span a range of different evolutionary histories, distributions and ecologies: blue-winged kookaburra (_Dacelo leachii_), brown-backed honeyeater (_Ramsayornis modestus_), brown falcon (_Falco berigora_), and pale headed rosella (_Platycercus adscitus_). We then mapped the extent of occurrence for each species (Fig. 5). To do this, we obtained observation data from the Atlas of Living Australia across the whole of Australia [using the \texttt{ALA4R R} package; @r453], spatially thinned the data to omit points within `r cs1.params.LST[[MODE]]$thin.distance/1000` km$^2$ of each other to ameliorate the effects of sampling bias [using the developmental version of the \texttt{spThin R} package; \url{www.github.com/mlammens/spThin}; @r454], and fit `r cs1.params.LST[[MODE]]$mcp.percent` % minimum convex polygons [using the \texttt{adehabitatHR R} package; @r455]. We used this method because it is trivially reproducible using freely available data.

We generated `r cs1.params.LST[[MODE]]$dp.number` demand points for each species (Fig. S2). To achieve this, for each species, we generated random points inside the species range and at each point extracted the principal component values at that location. We then fit hyperbox kernels to the distribution of these points to characterise the realised niche of each species [using a manually chosen bandwidth of `r cs1.params.LST[[MODE]]$dp.bandwidth[1]` and a `r cs1.params.LST[[MODE]]$dp.quantile` quantile to map the core parts of the species' niches; implemented in the \texttt{hypervolume R} package; @r231]. We then generated uniformly distributed points inside the species' kernels and estimated the density of the training points at the uniformly generated points. These uniformly distributed points and associated density estimated were used as demand point coordinates and weights (respectively).

We generated two solutions. The first solution was generated using `r cs1.params.LST[[MODE]]$amount.target*100` % amount-based targets for all species. The second solution was generated the same amount-based target with an additional `r cs1.params.LST[[MODE]]$space.target*100` % niche-based for each species.

### Results
Generally, the solution generated using amount targets preserved a representative sample of each the `r as.character(as.english(length(cs1.params.LST[[MODE]]$species.names)))` bird species' niches--with `r ifelse(length(amount.not.represented.species.names)>1, as.character(as.english(length(amount.represented.species.names))), 'one')` excpetion. Specifically, this solution sampled over `r cs1.params.LST[[MODE]]$space.target*100` % of the realised niche of `r parsed.representative.space.held.names`. Yet it failed to achieve this for `r parsed.not.representative.space.held.names`. This result suggests that prioritisations generated using conventional methods may preserve a representative sample of most species realised niches (Figure S1). However, despite this, there may yet be species for which only a small fraction of their realised niche is preserved. By explictly using space targets, conservation planners can generate prioritisations that are guaranteed to capture a representative sample of species' niches.

@@

# Acknowledgements
JOH is funded by an Australian Postgraduate Award (APA) scholarship. RAF has an Australian Research Council Future Fellowship. This work was supported by the Centre of Excellence for Environmental Decisions (CEED) and the Landscape Ecology and Conservation Group (LEC) at The University of Queensland.

# Data accessibility
The \texttt{raptr R} package can be downloaded from The Comphrehensive R Archive Network (\url{https://cran.r-project.org/}). All data, code, and results are stored in an online repository (\url{https://www.github.com/paleo13/raptr-manuscript}) to permit replication and validation of this study.

>>>>

1. An underlying aim in conservation planning is to maximise the long-term persistence of biodiversity. To fulfill this aim, the processes that sustain biodiversity must be conserved. Many biodiversity processes can be preserved by conserving a representative sample of each biodiversity feature (species or ecosystem). For example, representing ability for individuals to persist in a range of climatic conditions might be crucial for the long-term persistence of a species in a world where environmental change is accelerating. Yet current approaches overwhelmingly focus on securing a target proportion of the features' geographic range.

2. To address this issue, we developed the \\texttt{raptr: Representative and Adequate Prioritisation Toolkit in R} to guide reserve selection using explicit targets for representing biodiversity processes as well as area protection targets. Users set \"space targets\" to secure a representative sample of each feature within an attribute space that reflects a biodiversity process (eg. an attribute space expressing variation in genetic characteristics between individuals, or variation in climatic conditions between areas). Users also set \"amount targets\"--similar to conventional reserve selection methods--to secure a target proportion of the feature's geographic distribtuion. We demonstrated the functionality of this package using a simulated study and two case studies. In each study, we generated solutions that secured a proportion of the species' distributions using amount targets--representing conventional reserve selection--and compared them with solutions generated using both amount and space targets.

3. We show that explicitly considering biodiversity processes in reserve selection can substantially change the solution. The simulations suggest that including representativeness in conservation planning is important where biodiversity features have multi-model distributions in an attribute space. The case studies show that explicitly considering space-based targets for biodiversity processes can return solutions that could be far more effective in achieving long term biodiversity persistence and do not require a much greater area reserved than traditional approaches.

@@

<!--  title page -->
Short running title: Representative and adequate prioritisations \newline
Abstract word count: XXX / XXX  
Total word count: XXX / XXX  
Number of references: XXX  
Number of figures: XXX  
Number of tables: 0  
Number of textboxes: 0  
Number of equations: XXX  
\newpage

<!-- conservation and reserve selection -->
Perhaps the most fundamental aim of conservation is to maximise the long-term persistence of biodiversity [@r423; @r255]. To achieve this, conservation actions must preserve biodiversity patterns (eg. populations, species, ecosystems), but also crucially the processes that sustain them. One of the major tangible achievements of modern conservation has been the act of setting aside areas for preservation [@r440]. Reserve networks buffer species from gross threatening processes and set the stage for enhanced management interventions [@r465]. Since the resources available for conservation action are limited, protected area networks must be sited in places that satisfy conservation objectives for minimal cost [@r255]. However, despite the importance of biodiversity processes, they are only rarely used to guide conservation planning.

<!-- biodiversity processes are important -->
Prioritisations must capture the ecological and evolutionary processes that sustain biodiversity [@r3; @r255; @r9]. Ecological processes are important for biodiversity to persist over short time-scales. For instance, predator-prey processes can mediate ecosystem structure and function through top-down control [@r476]. Additionally, migratory processes allow many species to access suitable habitat throughout the year [r477]. On the other hand, evolutionary processes are important for biodiversity to persist over long time-scales. Adaptive evolutionary processes foster resilience against environmental change [@r3; @r43], and neutral evolutionary processes maintain genetic diversity [@r5; @r67; @r63]. In recent decades, a wealth of data relevant to biodiversity processes has become available [eg. bioclimatic, distribution, genetic, phylogenetic, and trait data; @r406; @r33; @r460; @r453]. This is sparked significant interest in using this data to make better conservation decisions [@r473; @r9; @r448; @r2; @r68; @r470; @r469; @r247; @r488; @r468; @r407; @r61; @r475].

<!-- current methods for preserving biodiversity processes -->
Today, the most widely used conservation planning tools focus on biodiversity patterns [\texttt{Marxan} and \texttt{Zonation}; @r22; @r429]. The cornerstone of these tools is that decision makers can set targets for a proportion of the geographic range of species, or ecosystems (termed features). To incorporate some information on biodiversity processes, current methods involve partitioning the features into different groups as a pre-processing step [@r9; @r448; @r247]. These groups are partitioned based on a proxy variable that relates to a biodiversity process of interest [@r473]. For instance, species ranges' have been partitioned according to habitat discontinuities to capture adaptive processes [@r2]. However, by partitioning features into different groups, it is not clear how well a given solution samples the original proxy variable, and ultimately, how well the solution captures the biodiversity process of interest. To overcome this limitation, Faith and Walker [-@r218] developed a decision support tool (\texttt{DIVERSITY}) to explicitly sample proxy variables to secure biodiversity processes. However, this tool is not widely used today, in part, because it cannot accommodate explicit targets on biodiversity patterns for multiple features.

<!-- aims of this paper -->
One of the key issues in reserve selection is the lack of unified a decision support tool that uses explicit and separate targets for biodiversity patterns and processes. To begin to fill this gap, we united the ideas underpinning \texttt{DIVERSITY} and \texttt{MARXAN} into new optimisation problems, and implemented them in the \texttt{raptr R} package. This \texttt{R} package provides decision makers with the tools to generate prioritisations that preserve both biodiversity patterns and processes. Here, we aim to the provide an in-depth understanding of its inner workings and explore its functionality.

# Methods
## PROBLEM FORMULATION
Biodiversity features are defined as the entities that the prioritisation is required to preserve (eg. species, ecosystems). Spatial attributes are defined as the variation across the feature' range that the prioritisation is required to sample. They can be intrinsic (eg. genetic or trait variation) or extrinsic (eg. environmental variation) to the feature. These attributes should be related to the biodiversity processes that the decision maker aims to preserve.

A set of attributes is conceptualised as a space (termed an "attribute space"). Each planning unit may be thought to exist as a single point inside a given space. For example, a decision maker may require a prioritisation that represents populations along climatic gradients. To achieve this, the decision maker might use an "climatic" attribute space with dimensions relating to mean annual temperature ($^{\circ}$C) and precipitation (mm). Any given combination of temperature and precipitation may be conceived as a point in this environmental space. By associating planning units with climatic data, they can be mapped from geographic space to this environmental attribute space. In addition to planning units, attribute spaces also contain demand points.

Demand points are designated by the decision maker to indicate regions of the attribute space that they wish to represent in the prioritisation (see below for discussion on how demand points can be generated for real-world data sets). The amount of variation in the attribute space that a solution secures depends on the distance between each demand point and each selected planning unit in the attribute space. The shorter the distances between the demand points and the planning units; the better the prioritisation is at securing the variation in the spatial attribute. To calculate the proportion of variation secured in a given solution, the distances are scaled by the distance between the demand points to their centroid (equivalent to calculating R$^2$ values for K-means clustering). In any attribute space there may exist points that are impossible (eg. mean annual rainfall -5 mm), or do not occur in the study area (eg. mean annual temperature 30$^{\circ}$C in Antarctica). Additionally, there may be some regions that are desirable for some features and undesirable for others (eg. conditions known to be outside the physiological tolerance of certain species). Thus a different set of demand points and weights are used for each attribute space and each feature. By using demand points in desirable regions of an attribute space for a given feature, the decision maker can ensure that prioritisations secure the feature in planning units with spatial attributes that are desirable for that feature.

To illustrate these concepts, consider the following example: we wish to develop a prioritisation for a single species that has four populations. However, we can only afford to preserve three of the four populations. To ensure that we preserve a representative sample, we want to want preserve populations inhabiting different environmental conditions. We obtained annual rainfall (mm) and temperature ($^{\circ}$C) data at the location of each population. We used this data to construct a two-dimensional environmental attribute space. Next, we generated demand points as equi-distant points inside this space. By comparing the distributions of demand points and populations in the attribute space, we can identify a suitable prioritisation (Fig. 1). We can see that a solution that prioritises both populations _A_ and _C_, effectively "doubles-up" on the same environmental characteristics and would contain in considerable redundancy. Instead, a more representative sample of the intra-specific variation could be captured by securing populations $A$, $B$, and $D$. This example demonstrates how attribute spaces and demand points relate to biodiversity processes, and can be used to guide reserve selection process .

The formulations used to express the reserve selection problem in the \texttt{raptr R} package are based on a combination of the \texttt{Marxan}, \texttt{DIVERSITY}, and the uncapacitated facility location problems [@r451]. Although the \texttt{raptr R} package utlises two novel formulations, for brevity, we will define the simpler formulation--referred to as the unreliable formulation--below and define the more complex version--the reliable formulation--in the Supporting Information S1. These formulations are named after the unreliable and reliable facility location formulations from which they are based upon [@r16]. The key difference between these two formulations is that the reliable formulation explicitly considers the probability that planning units are occupied when calculating the proportion of an attribute space sampled in a solution. All mathematical terms defined hereafter are described in Table S1. For convenience, the cardinality of sets will be denoted using the same symbol used to denote the variable.

@@

Targets are used to ensure that prioritisations adequately preserve each feature. Amount-based targets are used to ensure that the total amount of habitat preserved is sufficient. Let $T_f$ denote the expected amount of area that needs to be preserved for each feature $f \in F$. Space-based targets ensure that a representative sample of each feature is secured. Let $\tau_{fs}$ denote the space-based targets for feature $f \in F$ and attribute space $a \in A$. For convenience, all targets are expressed as proportions in the \texttt{R} package.

@@

\begin{align*}
T_s &= \text{amount target for feature $f$} \tag*{eqn 1a}\\
%
\tau_{sa} &= \parbox{25em}{representation target for feature $f$ in attribute space $a$ according to the corresponding demand points} \tag*{eqn 1b}\\
%
B &= \text{boundary length modifier (BLM): penalise fragmented solutions} \tag*{eqn 1c}\\
\end{align*}

@@

\begin{align*}
X_j
	&= \begin{cases}
		1, & \parbox{25em}{if planning unit $j$ is selected for conservation action} \tag*{eqn 2a} \\
		0, & \parbox{25em}{otherwise} \\
	\end{cases} \\
%
Y_{fsij} &= \begin{cases}
		1, & \parbox{25em}{if demand point $i$ is used to represent planning unit $j$ for feature $f$ in space $s$. } \tag*{eqn 2b} \\
		0, & \parbox{25em}{otherwise} \\
	\end{cases} \\
\end{align*}

@@

## OPTIMISATION
The unreliable formulation is non-linear. However, the non-linear components can be linearised using existing techniques. The expression $X_j X_k$ in (eqn 3a) can be linearised using methods described by Beyer et al. [-@r426]. Linearised versions of the problems can be solved using commercial exact algorithm solvers. The \texttt{raptr} R package provides functions to express conservation planning data as optimisation problems using linearised versions of the unreliable and reliable formulations (see Supporting Information S1 for formulation). These optimisation problems can then be solved to generate prioritisations using the commercial \texttt{Gurobi} software suite (\url{http://www.gurobi.com}). Presently, academics can obtain a license at no cost from the Gurobi website.

@@

## SIMULATION STUDY
### Methods
We simulated a hypothetical study area with planning units arranged in a $`r sqrt(sim.params.LST[[MODE]]$number.planning.units)` \times `r sqrt(sim.params.LST[[MODE]]$number.planning.units)`$ square grid (Fig. 2). We then simulated three species across this study area. The first species was simulated to represent a hyper-generalist--occurring in all planning units with equal probability (Fig. 2a). This species' distribution was based on a uniform distribution (eqn 8a). The second species was simulated to represent a species with an idealised distribution (Fig. 2b). It has a core range area, and is less likely to be found in areas that are distant from the core area. This species' distribution was simulated using the probability density function of a single multivariate normal distribution (eqn 8c). The third species represents a species with two distinct populations (Fig. 2c). This species' distribution was simulated to depict a bimodal distribution, based on the combination of the probability density functions of two multivariate normal distributions (eqn 8d). The probability that each species inhabited a given planning unit was calculated using the $X, Y$ coordinates of the planning unit and eqns 8a--8d. We used a geographic attribute space to showcase the behavior of the problem. For each species, demand points were generated by calculating the centroids of each planning units and their weights were set as the probability that the planning units were occupied.

@@

### Uniformly distributed species

The uniform species was simulated to occur with a constant probability of occupancy across the study area (Fig. 2a). The solution generated using `r sim.params.LST[[MODE]]$amount.target*100` % amount-based targets (Fig. 3a) selected `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units near the southern end of study area. This specific configuration is an artifact of the \texttt{Gurobi} solver used to solve the problem. In fact, for this species, all prioritisations containing `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units are optimal when considering only `r sim.params.LST[[MODE]]$amount.target*100` % amount targets. In the absence of criteria to guarantee representativeness, reserve selection methods may or may not return solutions that secure a representative sample of the features. In this particular case, the solution does not secure a representative sample of the species' range (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$space.held,2)` % sampled). Note that the proportion here is negative because it performs worse than simply siting a planning unit in the centroid of the species' demand points.

@@

### Normally distributed species

@@

By using `r sim.params.LST[[MODE]]$amount.target*100` % amount targets and `r sim.params.LST[[MODE]]$space.target*100` % geographic targets and a boundary length modifier parameter ($BLM = 1$), we obtained a solution that fulfills adequacy, representativeness and connectivity objectives (Fig. 3h). This solution contains `r filter(sim.spp.DF, species=='Normal\nspecies', prioritisation=='Amount & space\ntargets and BLM')$n` planning units. Similar to the corresponding solution for the uniform species (Fig. 3d), this solution contains more planning units than any other solutions for this species (cf. Figs. 3e--3g). The results for the normally distributed species suggest that the space targets can result in solutions that secure a more representative sample of the species'--even for species without significant spatial structure.

### Bimodally distributed species

@@

## CASE STUDY 1
### Methods
We investigated how space-based targets can be used in a multi-species planning context to generate a prioritisation that sufficiently preserves the realised niche for several species. By preserving the populations in suitable habitats with different environmental conditions, conservation planners can preserve the species' adaptive landscape and foster resilience against environmental change [@r63]. We selected Queensland, Australia as the study area. This region is ideal for exploring the potential of niche-based targets because it contains a broad range of different habitats. We obtained data for 19 bioclimatic variables across the region [at $30^{\prime \prime}$ resolution from \url{www.worldclim.org}; @r33] and subjected them to a principal components analysis (using ArcMap 10.3.1). We used the first `r as.character(as.english(cs1.params.LST[[MODE]]$n.pc.layers))` principal components (cumulatively explained `r round(cs1.pca.DF[[4]][cs1.params.LST[[MODE]]$n.pc.layers],1)` % of the total variation) to characterise the environmental variation across the study area (Fig. 4).

We selected four bird species that span a range of different evolutionary histories, distributions and ecologies: blue-winged kookaburra (_Dacelo leachii_), brown-backed honeyeater (_Ramsayornis modestus_), brown falcon (_Falco berigora_), and pale headed rosella (_Platycercus adscitus_). We then mapped the extent of occurrence for each species (Fig. 5). To do this, we obtained observation data from the Atlas of Living Australia across the whole of Australia [using the \texttt{ALA4R R} package; @r453], spatially thinned the data to omit points within `r cs1.params.LST[[MODE]]$thin.distance/1000` km$^2$ of each other to ameliorate the effects of sampling bias [using the developmental version of the \texttt{spThin R} package; \url{www.github.com/mlammens/spThin}; @r454], and fit `r cs1.params.LST[[MODE]]$mcp.percent` % minimum convex polygons [using the \texttt{adehabitatHR R} package; @r455]. We used this method because it is entirely reproducible using freely available data.

We generated `r cs1.params.LST[[MODE]]$dp.number` demand points for each species (Fig. S2). We used the following method to ensure that the demand points reflected the core parts of the species' realised niches. For each species, we generated random points inside the species range and at each point extracted the principal component values at that location. We then fit hyperbox kernels to the distribution of principal component values to characterise the realised niche of each species [using a manually chosen bandwidth of `r cs1.params.LST[[MODE]]$dp.bandwidth[1]` and a `r cs1.params.LST[[MODE]]$dp.quantile` quantile to map the core parts of the species' niches; implemented in the \texttt{hypervolume R} package; @r231]. We then generated uniformly distributed points inside the species' kernels and estimated the density of the training points at the uniformly generated points. These uniformly distributed points and associated density estimated were used as demand point coordinates and weights (respectively).

We generated two solutions. The first solution was generated using `r cs1.params.LST[[MODE]]$amount.target*100` % amount-based targets for all species. The second solution was generated the same amount-based target with an additional `r cs1.params.LST[[MODE]]$space.target*100` % niche-based target for each species.

### Results
Generally, the solution generated using amount targets preserved a representative sample of each the `r as.character(as.english(length(cs1.params.LST[[MODE]]$species.names)))` bird species' niches--with `r ifelse(length(amount.not.represented.species.names)>1, as.character(as.english(length(amount.represented.species.names))), 'one')` excpetion. Specifically, this solution sampled over `r cs1.params.LST[[MODE]]$space.target*100` % of the realised niche of `r parsed.representative.space.held.names`. Yet it failed to achieve this for `r parsed.not.representative.space.held.names`. This result demonstrates that though solution generated using conventional methods may preserve a representative sample of some features (Figure S1), only through the use of explicit targets can decision makers be sure that solutions secure a representative sample for all features.

@@

# Acknowledgements
JOH is funded by an Australian Postgraduate Award (APA) scholarship. RAF has an Australian Research Council Future Fellowship. This work was supported by the Centre of Biodiversity and Conservation Science (CBCS).

# Data accessibility
The \texttt{raptr R} package can be downloaded from The Comphrehensive R Archive Network (\url{https://cran.r-project.org/}). All data, code, and results are stored in an online repository (\url{https://www.github.com/jeffreyhanson/raptr-manuscript}) to permit replication and validation of this study.
